---
title: "01-eda"
output: html_document
---

### Load Libraries
```{r}
library(tidyverse)
library(janitor)
library(ggplot2)
library(ggthemes)
library(ggcorrplot)
```

### Load Data
```{r}
abalone <- read_csv("../data/raw/abalone.csv")
```

**NOTE**: Length, diameter, and height are measured in millimeters; whole weight, shucked weight, viscera weight, and shell weight are measured in grams.

## Section 1: Cleaning the Data

### Tidying Column Names
```{r}
abalone <- abalone %>% clean_names()
```

### Checking Data Structure
```{r}
dim(abalone)
glimpse(abalone)
colSums(is.na(abalone))
head(abalone)
summary(abalone)
```

### Investigating Height == 0
```{r}
abalone %>% filter(height == 0)
```

Could the 0 be a measurement or recording error? Both are infant abalone, but is it possible for an abalone to have 0 millimeters (mm) of height?

### Investigating Height > 0.3
```{r}
abalone %>% filter(height > 0.3)
```

**Height = 1.13:** So looking at the length, diameter and whole weight measurements being modest around 0.35 to 0.6, height being bigger than all three seems biologically impossible. The shell height should typically be smaller than what the length and diameter are. This suggests that a decimal was misplaced when recording the height of this abalone, and it possibly should be 0.113. 

**Height = 0.515**: This could likely be a case where the abalone is abnormally large. Length and diameter seem large in proportion around the 0.56 and 0.7 range. Whole weight is also quite large at 2.21. 

### Ratio
```{r}
abalone %>%
  mutate(height_diameter_ratio = height / diameter) %>%
  pull(height_diameter_ratio) %>%
  summary()
```

**Height = 0.515**: Ratio is around 0.91, which is about 2.7 times the mean of the height diameter ratio. While it's high, it doesn't seem impossible for an abalone to be of this height based on its adjacent physical measurements. This happens to be a thick-shelled, hefty abalone. 

**Height = 1.13:**: Ratio is around 3.18, which coincidentally, that is approximately the same value as the max height diameter ratio. 

### Handling Height Outliers
I began by removing the two rows with height == 0, as from a biological standpoint, this doesn't seem possible. I also decided to remove the row where height == 1.13, because our height-diameter ratio shows that it's an extreme outlier that could likely be due to a recording error. The typical ratio is around 0.3414, while the ratio for this abalone is around 3.18. Since we have over 4000 rows of data, removing these three will not have a major impact on our analysis. I will keep the row with height == 0.515 and flag it in case it creates concerns in the analysis.

### Removing Height Outliers
```{r}
abalone_clean <- abalone %>%
  filter(height != 0 & height != 1.13)

dim(abalone_clean)
```

## Section 2: Distributions

### Rings
```{r}
ggplot(abalone_clean, aes(x = rings)) +
  geom_histogram(binwidth = 1) +
  labs(title = "Distribution of Rings Among Abalones", x = "Number of Rings", y = "Count") +
  theme_minimal()
```

### Length
```{r}
ggplot(abalone_clean, aes(x = length)) +
  geom_histogram(binwidth = .05) +
  labs(title = "Distribution of Length Among Abalones", x = "Length (mm)", y = "Count") +
  theme_minimal()
```

While `Rings` showed a moderate right-skewed distribution, `Length` shows a left-skew distribution, gradually rising from 0.1 to about 0.6, then dropping sharply on the right side. This may reflect that most of the sampled abalone in this dataset were mature adults, and very few of them were young or small individuals.

### Diameter
```{r}
ggplot(abalone_clean, aes(x = diameter)) +
  geom_histogram(binwidth = .02) +
  labs(title = "Distribution of Diameter Among Abalones", x = "Diameter (mm)", y = "Count") +
  theme_minimal()
```

`Diameter` also has a left-skewed distribution, rising gradually and peaking at approximately 0.48 mm and dropping sharply off on the right side. Given the fact that `Length` and `Diameter` share a nearly identical distribution, this signals that these variables are highly correlated and may introduce issues of multicollinearity when modeling. Biologically, it makes sense that they would be correlated, as these measurements capture the overall shell size from different axes, and as abalones grow, their shell tends to scale proportionally in all directions rather than expanding in only one dimension. This shared pattern of growth is likely the main factor of the strong correlation between the two variables.

### Height
```{r}
ggplot(abalone_clean, aes(x = height)) +
  geom_histogram(binwidth = .01) +
  labs(title = "Distribution of Height Among Abalones", x = "Height (mm)", y = "Count") +
  theme_minimal()
```

The outlier that was mentioned earlier, with the abalone that had a height of 0.515 mm, is seen on the far right side of the graph.

```{r}
#Same graph without outlier
abalone_clean %>% filter(height < 0.3) %>%
  ggplot(aes(x = height)) + 
  geom_histogram(binwidth = 0.01) + 
  labs(title = "Distribution of Height Among Abalones (Excluding Outlier)", x = "Height (mm)", y = "Count") +
  theme_minimal()
```

If we take out the outlier for a moment, we can see that the distribution of `Height` is roughly symmetric. It peaks around the 0.15 mm range and gradually falls after the peak to about the 0.25 mm range. This distribution ranges from 0.01 to 0.25 mm. The retained outlier sits well outside this range, which could disproportionately inflate the mean and possibly affect a fitted regression line, pulling it outside the range of the bulk of the values. This will be assessed in the model diagnostics phase once a baseline model is fit and leverage and influential statistics are examined. 

### Weight Variables
```{r}
abalone_long <- abalone_clean %>%
  select(whole_weight, shucked_weight, viscera_weight, shell_weight) %>%
  pivot_longer(cols = everything(), names_to = "measurement", values_to = "value")

ggplot(abalone_long, aes(x = value)) +
  geom_histogram(binwidth = 0.05) +
  facet_wrap(~ measurement, scales = "free") +
  labs(
    title = "Distribution of Weight Measurements Among Abalones",
    x = "Weight (g)",
    y = "Count"
  ) +
  theme_minimal()
```

All four weight variables share a similar right-skew distribution. They seem to peak early, roughly between 0.15 to 0.3 grams for `shell_weight`, `shucked_weight` and `viscera_weight`, and they also have a long gradual tail stretching to the right. Because `whole_weight` encompasses a much larger range, its peak is slightly larger around 0.8 to 1.0 grams, but it still shares the same shape as the other weight variables. In a biological sense, weight is usually not symmetric in shape. An abalone cannot weigh less than 0 grams, so there is a sharp zero-bound on the left side of the data distribution. (Note: I double-checked all the weight variables to ensure that there was no abalone in the data that had a weight of 0 grams, as the graphs looked slightly misleading on that account) Most of the population weighs a small amount, and a small number of heavy abalone pull the tail of the distribution towards the right. Our target variable `Rings` also had a right-skewed distribution, which could imply that weight variables may have a more naturally linear relationship with rings than `length` / `diameter` / `height` do. This is only a hypothesis right now and will be checked in the model diagnostics phase.

### Sex (Categorical Factor)
```{r}
ggplot(abalone_clean, aes(x = sex)) +
  geom_bar() +
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.75) +
  labs(title = "Distribution of Sex Among Abalones", x = "Sex", y = "Count") +
  theme_minimal()
```

### Sex Percentages
```{r}
abalone_clean %>%
  count(sex) %>%
  mutate(percentage = round(n / sum(n) * 100, 1)) %>%
  knitr::kable(col.names = c("Sex", "Count", "Percentage (%)"))
```

Male (M) abalones have the highest distribution in this sample at a value of 1528, making up about 36.6% of the entire sample size. Infant (I) abalones are the second highest distribution at a value of 1340 (around 32.1% of the sample), and female (F) abalones make up the smallest distribution in the sample at a value of 1306 (31.3% of the sample). Since all three classes roughly make up about a third of the sample each, this is not a severe class imbalance situation. Something that is worth considering is that Infant (I) is technically a life stage rather than a biological sex like M/F. This distinction is worth acknowledging when encoding `sex` for modeling. Since I doesn't sit on the same conceptual axis as M or F, its coefficient in a linear model would essentially capture age ('youth') as much as it would capture 'sex', which needs to be addressed when interpreting model estimates.

## Section 3: Correlation 

### Numeric
```{r}
abalone_clean %>%
  select(length, diameter, height, whole_weight, shucked_weight, viscera_weight, shell_weight, rings) %>%
  cor()
```

Nearly every single one of the numeric predictors have high correlation with one another, with `length` and `diameter` being the highest at 0.9868. Across the whole dataset, we have a strong multicollinearity problem. The variable that is the most correlated with `rings` is `shell_weight` (0.628). Notably, the predictor-to-predictor correlations are all considerably stronger than each predictor's correlation to `rings`, even with `shell_weight`'s correlation with `rings` being below the average correlation (around 0.85) seen among the predictors themselves. What this signals about the data and the multicollinearity observed is that physical size cannot fully determine the age of an abalone. Two abalone of similar size could be of different ages depending on other factors, such as growth rate, environment or genetics. These confounding variables are not determined in the dataset. If a linear regression model is built on this data, coefficient estimates could be unstable or hard to interpret. Methods like Lasso or Ridge could be a potential approach to take given this context, as it will help stabilize coefficients in the presence of multicollinearity, with Lasso specifically able to shrink estimates to zero and eliminate less useful variables.

### Heatmap
```{r}
cor_matrix <- abalone_clean %>%
  select(length, diameter, height, whole_weight, shucked_weight, viscera_weight, shell_weight, rings) %>%
  cor()

ggcorrplot(cor_matrix, lab = TRUE, type = "lower") +
  labs(title = "Correlation Matrix of Abalone Measurements",
       x = "Variable 1",
       y = "Variable 2") +
  theme_minimal()
```

**Note the deeper red predictor-to-predictor blocks versus the lighter rings column**

## Section 4: Bivariate Distributions

### Shell Weight
```{r}
ggplot(abalone_clean, aes(x = shell_weight, y = rings)) + 
  geom_point(alpha = 0.1) +
  labs(
    title = "Number of Rings by Shell Weight in Abalones",
    x = "Shell Weight (g)",
    y = "Number of Rings"
  ) +
  theme_minimal()
```

There is a clear positive trend between `shell_weight` and `rings`, but it appears to be slightly curved rather than linear. `Rings` rises steeply along with `shell_weight` at first, but then the relationship flattens out beyond `shell_weight` = 0.4 to 0.45. The correlation between these two variables was 0.628, so a positive trend on the plot was expected. This curved, flattening pattern likely explains why the correlation came out moderate rather than high — Pearson correlation only measures the strength of a linear relationship, so it understates a real but non-linear association like this one./

The points become more widespread at this range as well; variance appears to increase as `shell_weight` increases. This may pose a problem when it comes to fitting a regression model, as linear regression assumes constant (not increasing) variance. Based on this first plot, we could be motivated to log transform either `rings` or `shell_weight` (or both). 

### Height
```{r}
ggplot(abalone_clean, aes(x = height, y = rings)) + 
  geom_point(alpha = 0.1) +
  labs(
    title = "Number of Rings by Height in Abalones",
    x = "Height (mm)",
    y = "Number of Rings"
  ) +
  theme_minimal()
```

The outlier observed earlier in the analysis can be detected on this plot, but it appears to be an unremarkable point as it has 10 rings, sitting right in the middle range of `rings`. Rather than being labeled as an outlier, this observation is a high leverage point, because it has a fairly normal y-value but an extreme x-value, sitting outside far from the main data cluster. What is interesting about this plot is that `height` has a relationship to `rings` that is opposite to the relationship that was first observed with `shell_weight` and `rings`. Rather than curving and flattening as `height` increases, it seems to steepen upwards around the 0.15 to 0.2 mm range. Variance in `rings` apears to increase as well as `height` increases, so again, this could raise a potential concern with heteroscedasticity as model and residual diagnostics are evaluated. 

### Length
```{r}
ggplot(abalone_clean, aes(x = length, y = rings)) + 
  geom_point(alpha = 0.1) +
  labs(
    title = "Number of Rings by Length in Abalones",
    x = "Length (mm)",
    y = "Number of Rings"
  ) +
  theme_minimal()
```

As `length` increases, `rings` increases as well, which was expected given their correlation of 0.556. The relationship shows a similar curved, flattening shape with increasing spread at higher values, consistent with the patterns already observed in `shell_weight` and `height`. This pattern is likely a reflection of `length`'s strong correlation with those variables.

### Diameter
```{r}
ggplot(abalone_clean, aes(x = diameter, y = rings)) + 
  geom_point(alpha = 0.1) +
  labs(
    title = "Number of Rings by Diameter in Abalones",
    x = "Diameter (mm)",
    y = "Number of Rings"
  ) +
  theme_minimal()
```

`Diameter`'s relationship to `rings` is nearly identical to `length`'s relationship to `rings`. The pattern and spread is highly similar, and this is not a shocking revelation given the fact that `length` and `diameter` are highly correlated with one another (r = 0.987). 

### Shucked Weight
```{r}
ggplot(abalone_clean, aes(x = shucked_weight, y = rings)) +
  geom_point(alpha = 0.1) +
  labs(
    title = "Number of Rings by Shucked Weight in Abalones",
    x = "Shucked Weight (g)",
    y = "Number of Rings"
  ) +
  theme_minimal()

abalone_clean %>% filter(shucked_weight > 1.0) %>% nrow()
```

As `shucked_weight` increases, variance of `rings` increases, similar to what was seen in other weight and size variables. However, with this specific relationship, there appears to be a decline at the higher range of `shucked_weight`, and the points look more sparse. The amount of abalone with a `shucked_weight` greater than 1.0 grams is 42, as confirmed with the code above. Given the fact that we have 4,174 total observations in our clean data, this is about 1% of the whole dataset. This decline is more plausibly attributed to sparse data at the high end, rather than a genuine biological phenomenon where older abalone (with more rings) would have lower shucked weight. All in all, `shucked_weight` has a genuine but noisy relationship with `rings` due to being the weakest correlated predictor (r = 0.42).

### Whole Weight
```{r}
ggplot(abalone_clean, aes(x = whole_weight, y = rings)) +
  geom_point(alpha = 0.1) +
  labs(
    title = "Number of Rings by Whole Weight in Abalones",
    x = "Weight (g)",
    y = "Number of Rings"
  ) +
  theme_minimal()
```

### Viscera Weight
```{r}
ggplot(abalone_clean, aes(x = viscera_weight, y = rings)) +
  geom_point(alpha = 0.1) +
  labs(
    title = "Number of Rings by Viscera Weight in Abalones",
    x = "Viscera Weight (g)",
    y = "Number of Rings"
  ) +
  theme_minimal()
```

Both `whole_weight` and `viscera_weight` show a similar pattern on their plots. They have a positive trend with `rings`, and it's not a perfectly linear relationship as it slightly curves then flattens. As both `whole_weight` and `viscera_weight` increase, variance in `rings` also increase. All of these findings are consistent with what was observed comparing `rings` and `shell_weight`, but it differs from the apparent decline observed with `shucked_weight`. The fact that neither `whole_weight` nor `viscera_weight` show an apparent decline on their plots further suggests that `shucked_weight`'s decline was due to the sparseness of data points in the higher range of `shucked_weight` rather than a real biological effect.

### Sex
```{r}
ggplot(abalone_clean, aes(x = sex, y = rings)) +
  geom_boxplot() +
  labs(
    title = "Number of Rings by Sex of Abalones",
    x = "Sex",
    y = "Number of Rings"
  ) +
  theme_minimal()
```

Infant (I) seems to have a lower median number of rings than male and female abalones (both sitting at a median of 10 rings). The box for Infant (I) sits lower on the y-axis compared to the other two groups. The M and F groups are also noticeably similar to one another, sharing a median value of 10 and IQR based on the size of their boxes. This could signal that biological sex past the infant stage is not strongly related to ring count. Something that is interesting to note is that the male and female groups have outliers stretching towards 25 to 29 rings. The infant abalone group also has some apparent outliers past 20 rings, despite it representing the youngest life stage of abalones. This raises the question of whether the I group in sex strictly represents infant abalone. This is not a question the dataset alone can answer, but it's worth noting when using sex as a predictor variable in a model.

### Bivariate Distributions Conclusion

This dataset contains predictor variables that are highly correlated with one another, presenting an issue of multicollinearity when building out a model. While the predictors are correlated with each other, they show moderate correlation to the target response variable (`rings`). Through the scatterplots, we've gathered that the relationships between `rings` and many of the predictors are non-linear and that `rings` increases in variance as many of the predictors increase in value. 

With our only categorical variable `sex`, it likely doesn't seem to be a strong predictor of `age` based on similarities between male and female abalone in terms of median and IQR. Infant (I) does, however, differ from male (M) and female (F) abalone, as it seems to have a lower median number of rings and a lower IQR. This is consistent with what we'd expect biologically - infants, being the youngest life stage, having a comparatively lower distribution of rings (and therefore age). 

Given the severe multicollinearity between predictor variables and the heteroscedasticity problem found in many of the scatterplots, a linear regression model may produce unstable, hard-to-interpret coefficients. Regularized regression approaches like Lasso and Ridge are natural candidates to address the multicollinearity directly, and the non-linear, heteroscedastic patterns observed also motivate testing non-parametric models such as a random forest, as it doesn't assume linearity and handles correlated predictors more efficiently.